---
title: "Diversitree_Likelihood"
author: "Sergei Tarasov"
date: "2025-06-26"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(diversitree)
library(plyr)
library(expm)
```


# Mk Model

```{r}
# Read in tree and Plot
phy <- readRDS('asr/tree.rds')
cols <- mapvalues(phy$tip.state, from = c(0, 1), to=c("grey", 'black'))
plot(phy, label.offset = 2, cex=1, no.margin = F)
tiplabels(pch = 15, col = cols, cex = 2, adj = 1.5)
nodelabels()
tiplabels()
edgelabels()
axisPhylo()
```



```{r}
# define Ln function 
lik.mk <- make.mk2(phy, phy$tip.state, strict=T)
# estimate Ln given pars
pars.mk <- c(.1, .2)
lik.mk.est <- lik.mk(pars.mk, root=ROOT.FLAT, intermediates = T)
print(lik.mk.est)
```

## Definitions of lik.mk() object

- **`init`**: Variable values at the **tip** of each branch. Note: it seems like there a bug for tips printing NAs. The computations do work. When Probs of two sister branches are combined at a node, it's an init variable of a downstream brahnch

- **`base`**: Variable values at the **base** of each branch. Note: they are normalized, see below.

- **`lq`**: The **compensation factor** for each branch.  
  Recall that the compensation factor is the **log of the sum** of the variables at the base of each branch.

- **`pij`**: Are present only in Mk model. They are direct probs after exponentiation.

- **`root.p`**: is the root prior.

- **`vals`**: is the variables at the root node, they are the same **`init`** for node 6.


## Manual calculations

### Along branches

Let's consider sp8 and sp7 tips (branches 4 and 5) and go downstream to see how it works.

```{r}
Q <- matrix(c(-0.1,0.1,
              0.2, -0.2), 2, 2, byrow=T)
print(Q)

# branch length
br5 <- phy$edge.length[5]
# Probability
P5 <- expm(Q*br5) %*% c(0,1)
# compensation factor for branch 5
lq5 <- log(sum(P5))
# Normalized Probs at the base of the branch 5
base5 <- P5/sum(P5)

print(lq5)
print(base5)

print(attr(lik.mk.est, "intermediates")$lq[5])
print(attr(lik.mk.est, "intermediates")$base[,5])

# Note, the following vars are equal
log(P5) == log(base5) + lq5


```

Branch 4.

```{r}
br4 <- phy$edge.length[4]
P4 <- expm(Q*br4) %*% c(1,0)
lq4 <- log(sum(P4))
base4 <- P4/sum(P4)

print(lq4)
print(base4)

print(attr(lik.mk.est, "intermediates")$lq[4])
print(attr(lik.mk.est, "intermediates")$base[,4])
```

Branch 9: the ancestor of Branch 4 and Branch 5.

```{r}

# init array for branch 9: combining two bases
init9 <- base5*base4

# branch 9 length: note id differs from ape
br9 <- phy$edge.length[3]
# Now Probs are condition on init9, note it's normalized
P9 <- expm(Q*br9) %*% init9
lq9 <- log(sum(P9))
base9 <- P9/sum(P9)

print(init9)
print(lq9)
print(base9)

print(attr(lik.mk.est, "intermediates")$init[,9])
print(attr(lik.mk.est, "intermediates")$lq[9])
print(attr(lik.mk.est, "intermediates")$base[,9])

```

### At the root

```{r}

# Root vals at node 6.
root.vals <- attr(lik.mk.est, "intermediates")$vals
root.pi <- c(0.5, 0.5)
LQ <- attr(lik.mk.est, "intermediates")$lq

# Likelihood Manual
log(sum(root.vals * root.pi)) + sum(LQ)

# Likelihood diversitree
print(lik.mk.est[1])


```



# BiSSE Model

```{r}
# Read in tree and Plot
phy <- readRDS('asr/tree.rds')
cols <- mapvalues(phy$tip.state, from = c(0, 1), to=c("grey", 'black'))
plot(phy, label.offset = 2, cex=1, no.margin = F)
tiplabels(pch = 15, col = cols, cex = 2, adj = 1.5)
nodelabels()
tiplabels()
edgelabels()
axisPhylo()
```

## Make Ln functions

```{r}
lik <- make.bisse(phy, phy$tip.state)
pars <- c(.1, .2, .03, .06, .01, .02)
lik.est <- lik(pars, root=ROOT.FLAT, intermediates=T)
print(lik.est)
```

Now, extracting various functions for Ln estimation from lik().

```{r}
e <- environment(lik)
# Ln for all branches f
all_branches <- e$all_branches
# root f
rootfunc <- e$rootfunc


eb <- environment(e$all_branches)
# Ln along single branch f
branches <- eb$branches
# Ln at a node f
initial.conditions <- eb$initial.conditions
```

### branches() and initial.conditions()

1. branches(): along a branch 

2. initial.conditions(): at node

```{r}

#------ branches (y, len, pars, t0, idx)
#
# in do.asr.marginal.R():
# branches(y.in, len[i], pars, depth[i], i)
# for depth[i], i, I herein use zeros but it may differ in td models

# Branch 5, sp8 (black)
br5 <- phy$edge.length[5]
y5 <- c(0,0,0, 1)
base5.raw <- branches(y5, br5, pars, 0, 0)

# first is lq, second is base
print(base5.raw)
attr(lik.est, "intermediates")$lq[5]
attr(lik.est, "intermediates")$base[,5]

# Branch 4, sp7 (white)
br4 <- phy$edge.length[4]
y4 <- c(0,0, 1,0)
base4.raw <- branches(y4, br4, pars, 0, 0)
print(base4.raw)


#------ initial.conditions(init, pars, t, idx)
#
# in do.asr.marginal.R(): 
# initial.conditions(branch.base[,children[j,]],  pars, depth[j], j)

# At node 9
base4_5 <- cbind(base4.raw[[2]], base5.raw[[2]])
init9 <- initial.conditions(base4_5, pars, 0, 0)

print(init9)
attr(lik.est, "intermediates")$init[,9]
```


### At root

```{r}
#
# rootfunc(res, pars, condition.surv, root, root.p, intermediates)
vals <- attr(lik.est, "intermediates")$vals
LQ <- attr(lik.est, "intermediates")$lq
res <- list(vals=vals, lq=LQ)

Ln <- rootfunc(res, pars, condition.surv=T, root=ROOT.FLAT, root.p=NULL, intermediates=F)

print(Ln)
print(lik.est[[1]])

```



